\section{\fourten}
\label{sec:410}

\fourten, Operating System Design and Implementation,
is an upper-level undergraduate class
designed for juniors and seniors in Computer Science
and Electrical and Computer Engineering.
For CS students it is one of five courses which fulfill
a ``systems programming'' degree requirement.
%
The prerequisite course is \twothirteen,
Introduction to Computer Systems,
% TODO CAMREADY
%a course which has been adopted at other
%universities in the U.S.\ and worldwide
adapted from Carnegie Mellon University's 15-213
\cite{sigcse01:CSaPP}.
This section discusses the course learning objectives,
outlines the programming projects,
and mentions some current limitations in the scope
of the course material.

The course has many learning objectives,
ranging from acquiring detailed factual knowledge about
hardware features
through practicing advanced cognitive processes
such as open-ended design.
%%
%Students study high-level concepts
%such as
%\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
%file-system internals,
%and log-based storage.
%%
%Emphasis is placed on acquiring information from ``primary sources,''
%including both manufacturer-provided hardware documentation
%and a non-textbook technical-literature reading assignment.
%
The course's distinguishing feature is that
students begin with a ``blank slate'' rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements from multiple sources
and must
choose their own module boundaries and inter-module conventions.
%
%Due to the foundational nature of kernel code,
The assignment design and grading encourage students to
think about corner cases, including resource exhaustion,
instead of being satisfied by
``the right basic idea''
implementations that handle only auspicious situations.
%
Finally, most relevant to this work,
students %gain substantial experience in
analyze and write lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread
%concurrency
and thread/interrupt concurrency.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Projects}

In the course of a semester, students work on five
programming assignments; the first two are individual,
and the remaining three are the products of two-person
teams.

\subsubsection{Stack Crawler}

The first project is a backtrace library: %``stack crawler'':
when invoked by a client program, it displays the
%program's
stack symbolically,
showing function names, strings, floating-point values, etc.
This project
%\shortversion{enables}{meets a variety of goals: it enables}
%students to review key process-model and
%language-runtime concepts from the
%prerequisite course;
%it
introduces students to our expectations about
design, analysis, and making choices;
%finally,
and
because C pointers are unsafe, it requires students
to consider robustness.
%
% and we READ THEIR CODE
% ^^ LOL, classic dave.
This project is built using a standard C tool chain
and debugged as students see fit (generally \x{printf()} and GDB).

\subsubsection{Device Drivers and Game}

The second project is a simple game, such as Hangman,
which runs without an underlying
operating system.
We provide students a small C library %with selected C library routines
(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()}),
who then write a device-driver library consisting of
console output (to a memory-mapped text-mode video display),
keyboard input (interrupt-driven),
and a countdown timer.
The game enables students to test their drivers.
%
% and the need to write enough code to get into trouble
%

This project and the remaining ones are written in
C with some x86-32 assembly,
which is then
compiled and linked into an ELF executable,
stored into a %1.44-megabyte 3.5-inch
floppy-disk image,
and booted via GRUB.
If the image is copied to a real floppy or CD,
%or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can boot on standard PC hardware.
However, students most often use
Simics \cite{simics} (described in \sect{\ref{sec:simics}}).

\subsubsection{User-space Thread Library}

The third project is implementing a 1:1 thread library for
user-space programs,
%essentially a stripped-down
%version of POSIX Pthreads.
similar to POSIX Pthreads.
The primary goal is
for students to experiment
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose.
%(there is substantial variation across groups).
They then write other synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.

A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
There is a semantic gap
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and \x{deschedule()}.
%(the system call which suspends execution of a thread).

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO CAMREADY: deanonymize
in a twelve-page document~\cite{kspec-anonymized}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

\subsubsection{\pebbles Kernel}

For the fourth project, students %two-student teams
produce a kernel which implements the
same specification as the reference kernel
they previously relied on.
They design and implement some approach to
synchronizing and blocking threads while
they are in kernel space,
%\shortversion{,}{\footnote{
%Student designs for their kernel's synchronization primitives vary widely.
%For example, ``top of the class'' students who went on to become teaching assistants have alternately used condition variables but not semaphores, and semaphores but not condition variables.
%},}
a simple round-robin scheduler,
basic virtual memory,
a program loader,
%various x86 exception handlers,
and code for setting up and tearing down
threads and processes
(re-using their game-project device drivers).
These ingredients are combined to yield
twenty-five system calls,
which we list in Table~\ref{tab:syscalls}.
%(life-cycle, thread management, memory management,
%console I/O, and miscellaneous),
%which we discuss further in Section~\ref{sec:pebbles}.}

\begin{table}[t]
	\center \footnotesize
	\begin{tabular}{|l|p{0.30\textwidth}|}
		\hline
		\bf Syscall name & \bf Summary \\
		\hline
		\multicolumn{2}{c}{\em Lifecycle management} \\
		\hline
		\x{fork} & Duplicates the invoking task, including all memory regions. \\
		\x{thread_fork} & Creates a thread in the current task.\\
		\x{exec} & Replaces the current program in the invoking task with a new one. \\
		\x{set_status} & Records exit status of current task. \\
		\x{vanish} & Ends execution of the calling thread. \\
		\x{wait} & Blocks execution until another task terminates; collects its exit status.\\
		\x{task_vanish}* & Causes all threads of a task to \x{vanish}. \\
		\hline
		\multicolumn{2}{c}{\em Thread management} \\
		\hline
		\x{gettid} & Returns the ID of the invoking thread. \\
		\x{yield} & Defers execution to a specified thread. \\
		\x{deschedule} & Blocks execution of invoking thread. \\
		\x{make_runnable} & Wakes another \x{deschedule}d thread. \\
		\x{get_ticks} & Gets timer tick count since bootup. \\
		\x{sleep} & Blocks for a given number of ticks. \\
		\x{swexn} & Registers a user-space function as a software exception handler.\\
		\hline
		\multicolumn{2}{c}{\em Memory management} \\
		\hline
		\x{new_pages} & Allocates a specified memory region. \\
		\x{remove_pages} & Deallocates same. \\
		\hline
		\multicolumn{2}{c}{\em Console I/O} \\
		\hline
		\x{getchar}* & Reads one character from kbd input. \\
		\x{readline} & Reads the next line from kbd input. \\
		\x{print} & Prints a memory buffer to the console. \\
		\x{set_term_color} & Sets color for future console output. \\
		\x{set_cursor_pos} & Sets the console cursor location. \\
		\x{get_cursor_pos} & Retrieves the console cursor location. \\
		\hline
		\multicolumn{2}{c}{\em Miscellaneous} \\
		\hline
		\x{ls} & Loads a buffer with the names of files stored in the RAM disk ``file system.'' \\
		\x{halt} & Ceases execution of the OS. \\
		\x{misbehave}* & Selects thread-scheduling policy. \\
		\hline
	\end{tabular}
	\caption{The \pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
	\label{tab:syscalls}
\end{table}


%%% Justify by mentioning average lines of code. -- bblum
%For most students in the class, this is the
%largest and most complicated software artifact they
%have produced.
%Because the test suite and the grading criteria
%emphasize robustness and preemptibility of
%kernel code,
%there are many cross-cutting concerns.
%%
%Because the students are responsible for ensuring
%the run-time invariants underlying all compiler-generated
%code in the system (kernel and user-space),
%they gain experience with debugging at both the
%algorithm level and the register/bit-field level.

{\bf Lifecycle.} The process lifecycle is loosely based on the
Unix \x{fork()}/\x{exec()}/\x{wait()} model.
We adopt the Mach~\cite{DBLP:conf/usenix/AccettaBBGRTY86}
distinction between ``tasks,''
which are resource containers,
and threads,
each of which executes within a single task.
This differs from %partitioning is a different approach from that of
Plan~9's \x{rfork()}~\cite{Pike90plan9} or Linux's \x{clone()},
%both of
which allow a continuum of sharing
%between identical threads and independent processes,
at the cost of implementation complexity. %which we wish to avoid.
%A task's address space can be edited through two system
%calls that add and delete contiguous regions of memory.
%Students are expected to implement zero-fill-on-demand,
%but copy-on-write and demand-loading of executables are
%optional.
%Threads can print text to and read lines of text from
%the simple console that students implemented at
%the start of the semester.

Widely regarded as the most difficult concurrency problem in the project
is that of coordinating a parent and a child task that ``simultaneously''
exit:
% invoke \x{vanish()}:
when a task completes,
live children and exited zombies must be handed off
to the task's parent or to the system's \x{init} process,
at a time when the task's parent may itself be
exiting;
meanwhile, threads in tasks that receive new children
may need to be awakened from the \x{wait()} system call.
Due to design constraints imposed by other parts of the kernel specification,
careless solutions %solutions which are not carefully designed
are prone to data races or deadlocks.

{\bf Threads.}
%Within a task,
Threads are created via
\x{thread_fork}, which closely
mirrors the specification of \x{fork()}:
the new thread begins with identical %an exact copy of the
register values,
%of the old thread, except that
%the return-value register, \x{\%eax}, contains
%zero;
excepting \x{\%eax}, containing zero;
in the old thread, \x{\%eax} holds the thread ID
of the new thread.
This minimal formulation defers all policy decisions
%(e.g., stack size, location, and contents)
to the
user-space thread library.
A thread may suspend its execution via the
\x{deschedule()} system call,
and a blocked thread can be reactivated via
\x{make_runnable()}.

Each thread
can register an exception handler
%has the ability to register a handler
%for hardware exceptions (illegal instruction,
%division by zero, etc.)
via \x{swexn()},
which specifies the addresses of handler code
and a handler stack.
%When a thread encounters an exception,
The kernel reflects exceptions to user space
by
de-registering the handler,
pushing the thread's register state onto the handler stack,
and invoking the handler code.
The handler
%has the ability to
can re-register itself
(or some other handler),
run arbitrary code,
and/or atomically adopt a new set of register
values.
%The \x{swexn()} facility is used to enable
\x{swexn()} enables
the thread library to detect %when a managed
thread crashes
%thread has crashed
and %, more importantly,
to enable user-space automatic stack growth.
%implementation of automatically-growing stack
%regions.
%Removing the auto-stack policy and mechanism from
%its traditional position in Unix-like kernels
%simplfies the kernel and
%front-loads policy questions to an earlier
%part of the course,
%before students are facing the substantial pressure
%of the kernel project.



\subsubsection{\pebbles Extension}

Students who complete the kernel project on time
then work on a kernel-extension project.
Past projects have included
a sound card driver,
a filesystem,
suspend-to-disk,
profiling,
and an in-kernel debugger.
Two recent, more aggressive, projects have been
paravirtualization
%so that their kernels can host guest kernels and
and multi-processor support.
%to their single-processor kernels.

In the multi-processor project, students
%extended their kernels to
scheduled multiple threads across up to eight cores.
We provided base code for determining the CPU count, booting the application processors, and sending inter-processor interrupts (IPIs).
% Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
% to implement moving threads between processors,
% and
% to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
Students must move threads between processors,
synchronize memory unmapping across multiple TLBs,
%and account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
and find a way to prevent concurrent data access apart from disabling local CPU interrupts.
%
%To allow struggling groups to make progress without solving every synchronization challenge,
To help struggling groups,
we allowed
%the use of
a ``big kernel lock'', and rewarded groups according to how little they used it.
Submissions were also graded according to
correctness of the scheduler and spinlocks,
stability on a suite of stress tests,
and efficient use of IPIs. %for efficient processor coordination.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Simics}
\label{sec:simics}

The main execution and debugging
platform used in the class is
Wind River Simics~\cite{simics}.
%originally developed by the Swedish Institute of
%Computer Science and
%currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for correct code %that behaves correctly
but also for kernels that accidentally ``abuse'' hardware.
%{but also for code that is ``incorrect within reason.''
%Occasionally student code manages to invoke an unsupported
%function of some hardware device or even to cause the
%simulator to crash, but this is quite rare, and in almost
%every case when the ``bad event'' is characterized it turns
%out that what the software requested from the hardware was
%a mistake.
%Meanwhile, inside its large envelope, Simics supports
%\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86
%hardware-defined
descriptor tables.
%Unlike some symbolic debuggers,
%Simics is scriptable in Python,
%and it is possible to interpose user-written measurement
%frameworks to evaluate proposed hardware
%features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
%\shortversion{}{Performance is very reasonable (Simics can simulate certain
%cases, such as a CPU halted awaiting a clock interrupt,
%much faster than real time).}
%\shortversion{}{A major advantage of using Simics over the QEMU emulator in particular
%is that QEMU only issues timer interrupts at basic-block boundaries,
%which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}
It also provides instruction and memory access tracing and checkpointing/rewinding, both of which \landslide employs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Limitations}

%At present the course accepts the following limitations
%or core assumptions.

\textbf{32-bit x86.}
%\subsubsection{x86}
%\shortversion{}{Students write code that runs on one specific
%hardware platform.}
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them
test their code.
%boot
%up their code, type commands, and observe the results.
%\violence{}{Also, unlike many System-on-a-Chip platforms,
%detailed hardware descriptions are readily available.}
%% rather than being proprietary or "documented"
%% by labyrinthine Linux source code
%\shortversion{}{Finally, the platform is fairly stable---with only a
%little care, it is possible to write a compliant
%kernel which runs on anything newer than a 486DX processor,
%i.e., almost any PC hardware which is likely to be in
%working condition.}
%
%\subsubsection{32-bit}
We target the 32-bit
subset of x86, for multiple reasons.
%%% I'm not sure I believe this is an advantage. Instead of having
%%% to verify if each register is correct, students now have to verify
%%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
%First, while the small number of registers is
%frustrating for compiler writers,
%\shortversion{the}{from the point of view of OS students, the}
%existence of fewer registers means less work
%when it is necessary to verify that each register's
%value is correct.
%Likewise,
The calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater
insight.
%\shortversion{insight.}{insight. Finally, since it is impossible for an x86 processor
%to be in 64-bit mode with virtual memory off,
%students would need to be coached through the
%32-bit--to--64-bit transition or would need
%to be given a template virtual-memory implementation.}
%All in all, at present we are satisifed with
%the 32-bit choice.

\textbf{Memory model.}
At present we allow students
to reason about concurrency based on a
sequentially-consistent view of
% TODO: add a citation?
memory.
%which is consistent with
%existing x86 hardware~\cite{SewellSOZNM:x86tso-cacm10}
%but simplistic compared to most other
%architectures.
Lecture material educates students about
%the general topic of
out-of-order memory systems
and fencing,
but at present the hands-on experience does not
reinforce that material.

\textbf{Reduced breadth of implementation.}
One cost of programming to the complexities of
real hardware is that
within %the scope of
a single semester,
students do not %\violence{not}{not have the opportunity to}
implement common features of many kernels,
such as a file system, interprocess communication,
a real-time scheduler,
or networking.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Grading}
\label{sec:grading}

Thread libraries and kernels
are graded through a combination
of three methods:
performance on a test suite,
compliance with ``hurdle criteria,''
and detailed code review by a member of the course staff.

\textbf{Automated tests.}
%While developing their code,
Students will uncover both bugs
and design misunderstandings by running test code we
provide.
%Some tests are simple, exercising one feature or condition
%(e.g., allocating and freeing a large amount of memory),
%while other tests exercise a mix of kernel features
%simultaneously (creating and destroying processes,
%some of which do I/O while others handle or are killed
%by various exceptions) over an extended period of time.
The test suite is a number of user-space programs
which test
%functionality including
basic thread and process lifecycles, robustness
against corner-case security vulnerabilities, and long-term stability
under large stress tests.
%
%When students submit their thread libraries and kernels,
%we use an automated testing harness to
%run a superset of the public test suite.
%
Of course, due to the nondeterministic nature of many bugs in student code,
these tests can provide only a rough estimate of a submission's correctness,
rather than verifying it with certainty.
(This hit-or-miss nature motivated our work on \landslide.)
%Nevertheless, performance on these tests constitutes 30\% of the kernel project's grade.

\textbf{Hurdle completion criteria.}
Because of the ambitious learning objectives,
we employ a very detailed notion of completeness
for the thread library and kernel projects.
The ``hurdle'' model is based on the following observations:

\begin{itemize}
	\item Students learn more from bugs
		that are harder to find.
		In extreme cases, %code may do
		%almost everything right, but the last bug may
		subtle bugs may
		demonstrate a deep misunderstanding which requires a re-design.

	\item %Because concurrency bugs are triggered rarely,
		Finding rare concurrency bugs requires code capable of running
		for a long period without crashing.

	\item Robust code often requires a different structure
		than a ``usually works'' solution.
		It is easy to %write task-exit code which
		implement \x{vanish()} by
		allocating memory. %to notify the parent task.
		%parent that it has exited.
		However, such a design suffers from ``terminal irony'':
		a program wishing to exit,
		which would free up memory,
		cannot do so if memory is exhausted.
\end{itemize}

%For these reasons, \shortversion{much}{we believe much} of the learning
%associated with the kernel project necessarily occurs
%at the end of the experience,
%so it is important for students to finish the process.
% [suppressed: bad to build P4 on a shaky P3]
%\shortversion{It}{We believe it} would be particularly
It would be detrimental to
assign partial credit to a large body of non-working
code.
Instead, we require student thread libraries and kernels to pass a
specified percentage of tests
%each segment of the test suite
to be considered complete.
Furthermore,
because the test suite cannot detect certain
structural shortcuts (e.g., ``fixing'' a concurrency
problem by disabling interrupts during \textit{all}
kernel execution),
we require students to certify, in writing, that
their code possesses certain properties.

The thread library hurdle requires students to pass a suite of five test cases,
which test mutex locking/unlocking and thread creation, exit, and joining in moderately stressful ways.
%
A hurdle-passing kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain distasteful x86-specific code structures.

%Students who pass the completeness hurdle submit their kernels
%and can move on to the kernel-extension project.
%\shortversion{Students}{Meanwhile, students} who were unable to meet the
%hurdle criteria on time receive additional time
%to finish up and improve their code quality.
%%---this
%%requirement is based on our belief that important
%%learning experiences are obtained by solving the
%%hard problems,
%%which turn up at the end of the kernel implementation.
%The hurdle criteria constitute 50\% of the kernel project's grade.
%

\textbf{Code review and interview.}
Finally, both to make up for the tests' incompleteness and to provide personalized human feedback,
each student submission is manually graded by a member of course staff.
The instructors and teaching assistants print each group's submitted code and mark it with red ink, auditing it for style, architectural issues, races, and preemptibility.
Furthermore, after the kernel project, each group meets with their grader for a ``kernel interview'', essentially a debriefing for correcting any misconceptions.
% TODO: and what for p2?
Staff discretionary points constitute 20\% of the kernel project's grade.

