%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass[10pt,times,numbers]{sigplanconf}
\usepackage[1stsubmission]{oopsla2016}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

% template-provided usepackages
\usepackage{amsmath}
% custom ones copied from pldi template
\usepackage{url}
\usepackage{enumitem}
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

% ???
%\newcommand{\cL}{{\cal L}}

\begin{document}

% TODO CAMREADY: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
\newcommand\landslide{\textsc{Landslide}}
\newcommand\quicksand{\textsc{Quicksand}}
\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1} 
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{orange}{RGB}{192,96,0}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}
\definecolor{commentblue}{RGB}{0,0,192}

\newcommand\numthrlibs{79}
\newcommand\numpintoses{78} % 'di' is basecode. durr
\newcommand\numstudence{157} % total pintoses plus p2s

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{OOPSLA '16}{November 2--4, 2016, Amsterdam, the Netherlands}
\copyrightyear{2016}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Stateless Model Checking with Data-Race Preemption Points}

\authorinfo{Ben Blum}{Carnegie Mellon University}{bblum@cs.cmu.edu}
\authorinfo{Garth Gibson}{Carnegie Mellon University}{garth@cs.cmu.edu}
%\authorinfo{\vspace{24pt}}{}{}

\maketitle

\begin{abstract}
Stateless model checking is a powerful technique for testing concurrent programs,
but suffers from exponential state space explosion when the test input parameters are too large.
Several reduction techniques can mitigate this explosion,
but even after pruning equivalent interleavings, the state space size
%for a fixed set of preemption points
is often intractable.
Most prior tools are limited to preempting only on synchronization APIs,
which reduces the space further, but can miss unsynchronized thread communication bugs.
%
Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
It avoids concerns of state space size, but is prone to false positives:
spurious reports of access pairs that cannot be reordered to produce a bug.

We present an {\em Iterative Deepening} framework for stateless model checking,
which manages the exploration of many state spaces using different preemption points.
It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
and it incorporates data-race analysis to add new preemption points on the fly.
Preempting threads during a data race's instructions
%These preemption points allow us to check whether each data-race can produce an observable failure, and
can automatically classify the race as buggy or benign,
and uncovers new bugs not reachable by prior model checkers.
%
It also enables full verification of all possible schedules when every data race is verified as benign within the CPU budget.
In our evaluation, Iterative Deepening
%with data-race preemption points
found 1.8x as many bugs and verified 4.3x as many tests compared to stateless model checking without data-race preemption points.
%Our evaluation shows this technique is
%more effective than single-state-space model checking
%%both at finding more bugs and at completing more state spaces when no bug exists.
%both at finding the same bugs faster and at finding new bugs entirely.
\end{abstract}

\category{D.2.4}{Software Engineering}{Software/Program Verification}
\category{D.2.5}{Software Engineering}{Testing and Debugging}
%
%% general terms are not compulsory anymore,
%% you may leave them out
%\terms
%term1, term2
%
%\keywords
%keyword1, keyword2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{intro}
\input{overview}
\input{design}
\input{implementation}
\input{soundness}
\input{evaluation}
\input{warpzone}
\input{related}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:conclusion}

%We are great. Accept our paper.

We have presented Iterative Deepening and \quicksand, a new technique and tool for automating the choice of preemption points (PPs) during stateless model checking.
% , and \quicksand, a tool which implements this technique tailored to the \landslide~model checker.
\quicksand~incorporates data-race analysis to create new PPs tailored specifically to the program under test,
and automatically finds state spaces that are appropriately sized to complete in a given CPU budget.
%and manages multiple model checker instances to test many different PP subsets in a given CPU budget, even when the full state space of all PPs would be computationally intractable.

We achieve better bug-finding results than either single-pass data-race detection or single-state-space model checking alone,
%We dramatically reduce false-positive data race reports,
%%by verifying the absence of bugs in their associated state spaces
%and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
%Compared to prior work,
%, we find bugs faster in smaller subset state spaces, %% not true, or at least, not by very much
%when the maximal state space is too large to explore
finding new bugs with data-race PPs that could not be exposed by preempting only on synchronization APIs.
By using \numstudence~student projects as our test suite, we also show the potential benefit as a debugging platform in educational settings.
Moreover, when all data-race PPs can be fully tested within the CPU budget, we provide a verification as strong as if every single instruction had been used as a PP.

%Although \landslide~itself is not publicly available due to its dependency on the non-free simulator \simics, \quicksand~is open-source and its interface can be adapted to fit any similar tool.
\quicksand~is open-source and its interface can be adapted to fit any tool similar to \landslide.
We have also posted our evaluation's data set. %and logs.
They are available at:

% TODO CAMREADY: Make data set, at least, public.

{\em github links removed for double-blind review}
%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put QS in its own repository
%
%\url{https://github.com/bblum/landslide} % POST-SUBMISSION TODO: put final logs somewhere
% TODO CAMREADY: Don't forget to anonymize the names of all your grupos.

%\section{Acknowledgements}

% TODO CAMREADY
%{\em removed for double-blind review}
%Many thanks to Vaishaal Shankar, Haryadi Gunawi, and David A. Eckhardt for generously providing student implementations from Berkeley's, U. of chicago's, and CMU's OS classes respectively.
%Thanks to Wind River for the use of their simulator \simics.
%Thanks to
%Ji\v{r}\'{i} \v{S}im\v{s}a, Joshua Wise, Jean Yang, Brandon Lucia, Haryadi Gunawi, John Wilkes, and
%the anonymous PLDI reviewers for their helpful comments.
%This work was supported in part by
%the U.S. Army Research Office under grant number W911NF0910273.


% TODO: proofs->appeendix
\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{citations}{}

\end{document}
